---
title: "Linear models"
author: "Richard J. Telford"
date: "May 26, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Linear Modelling Exercises

1. Import the penguin data from the `palmerpenguin` package
````{r loading library}
library(tidyverse)
library(palmerpenguins)
penguins.df <- penguins

````

2. Do big birds have big beaks? Fit a linear model between bill length and body mass, and interpret the coefficients.
  Yes, because the p-value is significant. 
````{r lm }

fit1.lm <- lm(bill_length_mm~body_mass_g, data = penguins.df)
summary(fit1.lm)

````

````{r plot lm}
p1 <- penguins.df %>% 
  ggplot(aes(x = body_mass_g, y =bill_length_mm)) + 
  geom_point() + 
  theme_bw() + 
  labs(x = "Body mass (g)", y = "Bill length (mm)") + 
  geom_smooth(method = lm, color = "blue", formula = y~x, se = FALSE)
  
p1  

````

3. Examine the diagnostics plots. Is everything OK?
    Residual VS fitted: should not have a clear pattern - looks ok 
    Normal Q-Q: should follow the dashed line - looks ok 
    Scale-location: should see a fairly horizontal line with spread points - looks ok 
    Residual vs leverage: If you have points placed outside the dashed lines, they have high     Cook's distance. This means that they are influential outliers. Removing them will change     the results of the regression

````{r diagnostic plot}
par(mfrow=c(2,2)) # to get all four graphs in the same plot 
plot(fit1.lm) 

````

4. Calculate the residual sum of squares (find sum the squared difference between the estimates calculated from the coefficients and the observed values).

```{r RSS}
anova(fit1.lm)

##to do it by hand 
sum((yi-(b0+b1xi))^2) #formula for calculating RSS 

rss <- sum((penguins.df$bill_length_mm - #gives the observed value 
              ((2.69*10^1)+((4.051*10^-3)*penguins.df$body_mass_g)))^2, na.rm=TRUE)
```` 

5. Recalculate the Residual sum of squares using different values for the coefficients (in the formula yi = Beta + Beta xi + εi).

```{r RSS continued}
rss2 <- sum((penguins.df$bill_length_mm - #gives the observed value 
              ((4.87*10^1)+((2.92*10^-3)*penguins.df$body_mass_g)))^2, na.rm=TRUE)
rss2

````

6. Is the new RSS larger or smaller?

```{r}
The RSS gets larger, regardless of how I change the numbers. This shows that the model is a good fit because you want the RSS to be as low as possible. 

````

7. Estimate the slope using covariance and variance.

```{r slope calculation}
#covariance and variance 
penguins.df %>% 
  summarise(
          co_var = cov(penguins.df$bill_length_mm, penguins.df$body_mass_g, use = "pairwise.complete.obs"), 
          variance_bill_len = var(bill_length_mm, na.rm = TRUE), 
          variance_body_mass = var(body_mass_g, na.rm = TRUE))


slope <- ( cov(penguins.df$bill_length_mm, penguins.df$body_mass_g, use = "pairwise.complete.obs") ) / (sqrt(((var(penguins.df$body_mass_g, na.rm = TRUE))^2)*((var(body_mass_g, na.rm = TRUE))^2)))


2605.592/sqrt((643131.1^2)*(643131.1^2))

````

8. Do you get the same result?

9. Add species to the model as a predictor.

```{r add species as predictor}
fit2.lm <- lm(bill_length_mm~body_mass_g*species, data = penguins.df)

p2 <- penguins.df %>% 
  ggplot(aes(x = body_mass_g, y = bill_length_mm, color = species)) + 
  geom_point() + 
  theme_bw() + 
  labs(x = "Body mass (g)", y = "Bill length (mm)") +
  geom_smooth(method = lm, color = "blue", formula = y~x, se = FALSE)
p2  

summary(fit2.lm)

````

10. Does species improve the model?
  Yes, for this model the R-squared is higher and therefore more variance is explained by the model. 


11. How should the coefficients be interpreted?
  
  intercept: the bill length of Adelie penguins 
  body_mass_g
  speciesChinstrap: 
  speciesGentoo
  body_mass_g:speciesChinstrap
  body_mass_g:speciesGentoo 

