---
title: "Bio302 Practical 1 Exploratory Data Analysis"
author: "Richard J. Telford"
date: "May 25, 2016"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1 Peguins

Load the `palmerpenguins` package and see the `penguins` data.

````{r loading data}
library(palmerpenguins)
penguins.df <- penguins

```

Find the range, mean, median and variance of variable `bill_length_mm`.

````{r}
penguins.df %>% 
  summarise(mean = mean(bill_length_mm, na.rm = TRUE), 
                   median = median(bill_length_mm, na.rm = TRUE),
                   range = range(bill_length_mm, na.rm = TRUE), 
                   variance = var(bill_length_mm, na.rm = TRUE))
```

What is the standard error of the mean of `bill_length_mm`.

````{r standard error}
bill_len_SD <- sd(penguins.df$bill_length_mm, na.rm = TRUE)
bill_len_SD
bill_len_SE <- bill_len_SD/sqrt(sum(!is.na(penguins.df$bill_length_mm)))
bill_len_SE
```

Plot a histogram of variable `bill_length_mm` and mark the position of the mean and median.
````{r histogram, warning=FALSE, message=FALSE}
bill_length_histogram <- penguins.df %>% ggplot(aes(x = bill_length_mm)) +
  geom_histogram(fill = "grey") + 
  theme_bw() + 
  labs(title = "Bill length", x = "Bill length (mm)", y = "Count") +
  geom_vline(xintercept = mean(penguins.df$bill_length_mm, na.rm = TRUE), linetype = 1, color = "red")+
  geom_vline(xintercept = median(penguins.df$bill_length_mm, na.rm = TRUE), linetype = 1, color = "blue")
bill_length_histogram
```



Plot `bill_length_mm` against `bill_depth_mm`
There are three levels in factor `species` Plot `bill_length_mm` against `bill_depth_mm`, coloured by `species`

````{r plot bill length against bill depth, warning=FALSE}
plot <- penguins.df %>% ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point() + 
  labs(x = "Bill length (mm)", y = "Bill depth (mm)") 
plot
```

Find the covariance and correlation of `bill_length_mm` against `bill_depth_mm`.

````{r covariance and correlation}

#covariance 
cov(penguins.df$bill_length_mm, penguins.df$bill_depth_mm, use = "pairwise.complete.obs")

#correlation
cor(penguins.df$bill_length_mm, penguins.df$bill_depth_mm, use = "pairwise.complete.obs")

```

Is the correlation significant? 
  Yes, the p-value is less than 0.05 and is therefore significant.
  
````{r}
cor.test(x = penguins.df$bill_length_mm, y = penguins.df$bill_depth_mm, use = "pairwise.complete.obs")

````

How should the p-value be interpreted?
  When the p-value is significant, the null hypothesis is rejected. In this case this means that if true correlation was zero, it would be unlikely to get such a high correlation. 


Is the correlation between `bill_length_mm` against `bill_depth_mm` significant for any of the groups?

```{r}
penguinsA <- filter(penguins.df, species == "Adelie")
cor.test(x=penguinsA$bill_length_mm, y=penguinsA$bill_depth_mm, use = "pairwise.complete.obs")

penguinsC <- filter(penguins.df, species == "Chinstrap")
cor.test(x=penguinsC$bill_length_mm, y=penguinsC$bill_depth_mm, use = "pairwise.complete.obs")

penguinsG <- filter(penguins.df, species == "Gentoo")
cor.test(x=penguinsG$bill_length_mm, y=penguinsG$bill_depth_mm, use = "pairwise.complete.obs")

##Alternatively (a nicer code - copy this and do it for each species)
cor.test(~bill_length_mm + bill_depth_mm, data = penguins.df, use = "pairwise.complete.obs", subset = species == "Gentoo")

penguins.df %>% 
  group_by(species) %>% 
  summarise(cor = broom::glance(cor.test(bill_length_mm, bill_depth_mm, use = "pairwise.complete.obs" )))
````

This is a sub-group analysis, what are the dangers of this type of analysis?
  There are several limitations when it comes to sub-group analysis. 
    1. false positives due to multiple comparisons 
    2. false negatives due to inadequate power 
    3. small number of outcome events within each subgroup, whilst performing               multiple subgroup analyses makes it much more likely any effect suggested is         actually due to chance. If conclusions are made that a treatment is effective        based on subgroup analysis alone, those conclusions may well be misleading. 

# Part 2 not penguins

Install the `datasauRus` package with `install.packages("datasauRus")` and load it with `library`.

````{r data import}
library(datasauRus)

datasaurus.df <- datasaurus_dozen
`````

Dataset `datasaurus_dozen` has `x` and `y` variables and a dataset name. 
For each dataset, calculate the mean and standard deviation of x and y and the correlationn between x and y.

````{r analysis }
datasaurus.df %>% 
  group_by(dataset) %>% 
  summarise(mean_x = mean(x, na.rm = TRUE), 
            sd_x = sd(x, na.rm = TRUE), 
            mean_y = mean(y, na.rm = TRUE), 
            sd_y = sd(y, na.rm = TRUE), 
            cor_x_y = cor(datasaurus.df$x, datasaurus.df$y, use = "pairwise.complete.obs"))



````

datasaurus_dozen plot 

````{r plot}
datasaurus_plot <- datasaurus.df %>% 
  ggplot(aes(x = x, y = y, color = dataset)) + 
  geom_point() +
  theme_void() + 
  theme(legend.position = "none") + #remove the legend 
  facet_wrap(~dataset, ncol=3) #to make one plot per categorical variable 
datasaurus_plot
````




