---
title: "Practical 3"
author: "Richard J. Telford"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## GLS

With the built-in `iris` data.

````{r importing data, message = FALSE}
library(tidyverse)
iris.df <- iris
````

Make a plot to show how `Petal.Length` varies between species.


````{r plotting petal length}
petal_length_plot <- iris.df %>% 
  ggplot(aes(x = Species, y = Petal.Length, fill = Species)) + 
  geom_boxplot() + 
  labs(x = "Species", y = "Petal Length")
petal_length_plot
````

Find the variance of `Petal.Length` for each species.

````{r variance petal length}

iris.df %>% 
  group_by(Species) %>% 
  summarise(variance_petal_length = var(Petal.Length))


````

Fit an anova using `lm` between `Petal.Length` and species and examine the diagnostic plots.

````{r making a model}
fit1.lm <- lm(Petal.Length~Species, data = iris.df)
par(mfrow=c(2,2))
plot(fit1.lm) #bc of how the 3rd diagnostic plot looks, we can use gls for this dataset

summary(fit1.lm)

````

Fit a `gls` for the same model. Have the coefficients changed? (remember to load ´nlme´ package): No, the coefficients are the same. 

````{r gls , message = FALSE}
library(nlme)
fit1.gls <- gls(Petal.Length~Species, data = iris.df)
summary(fit1.gls)

````

Fit a `gls` for the same model but allow the variance to be different for each species by adding `varIdent` to the `weights` argument of the model. 

````{r }
fit2.gls <- gls(Petal.Length~Species, data = iris.df, 
                weights = varIdent(form = ~+1|Species ))
summary(fit2.gls)
````

Use `AIC` to test if this is a better model.

````{r AIC}
anova(fit1.gls, fit2.gls)

````

