---
title: "Spatial and temporal structure"
author: "Richard J. Telford"
date: "June 1, 2018"
output: html_document
---

```{r setup, include=FALSE, message= FALSE}

library(tidyverse)
library(nlme)
```

1) Simulate a 100-observation autocorrelated timeseries with `arima.sim`, with a first order autoregressive coefficient of 0.5. Also make a time vector of `1:100`

````{r}
x <- arima.sim(list(order = c(1, 0, 0), ar = 0.5), n = 100) #1, 0, 0 says it's a first order - write 2, 0, 0 if you want a second order autoreg. 
time <- c(1:100)

`````

2) Plot the data. 

````{r}

plot(x, col = "skyblue")

````

3) Regress the timeseries against time with an OLS model. Does the model appear to be statistically significant? No - the p-value is > 0.05 

````{r}

mod1 <- lm(x~time)
summary(mod1)
anova(mod1)

````

4) Plot the model diagnostics, including an `acf` and `pacf` of the residuals.


````{r}

plot(mod1)
acf(resid(mod1))
pacf(resid(mod1))

````


5) Use the Durbin-Watson test to test the residuals for autocorrelation.

````{r, message = FALSE}
library(lmtest)

dwtest(mod1)

````


6) Fit a gls with an appropriate correlation structure. Is this a better model? How have the p-value and effect size changed?

````{r}
fit1.gls<- gls(x~time)
summary(fit1.gls)
fit2.gls <- gls(x~time, corr = corAR1())
summary(fit2.gls)

anova(fit1.gls, fit2.gls)
````


7) Repeat the above 1000 times and find how autocorrelation affects the distribution of the p-value and the effect size.

````{r, message=FALSE}
library(broom)
library(broom.mixed)

rerun(10, fit2.gls) %>% 
  map_dfr(glance)

##with lm model 
rerun(1000, {x <- arima.sim(list(order = c(1, 0, 0), ar = 0.5), n = 100)
time <- c(1:100)
lm(x~time)}) %>% 
   map_dfr(glance) %>% 
  ggplot(aes(x = p.value)) + 
  geom_histogram()

##with gls model 
rerun(1000, {x <- arima.sim(list(order = c(1, 0, 0), ar = 0.5), n = 100)
time <- c(1:100)
gls(x~time, corr = corAR1())}) %>% 
   map_dfr(glance) %>% 
  ggplot(aes(x = p.value)) + 
  geom_histogram()

````

## Real data

1) The built-in dataset LakeHuron has annual lake level data from 1885 to 1972
Load the data with the command data(LakeHuron)

````{r}
data("LakeHuron")
view(LakeHuron)
lakehuron.df <- LakeHuron

````


2) Plot the data.

````{r}

plot(lakehuron.df, col = "skyblue")

````

3) Regress the LakeHuron lake level against year using a linear model. Is there a significant trend?

````{r}

year <- 
mod2 <- lm(LakeHuron~time(LakeHuron))
summary(mod2)

````

4) Plot the autocorrelation and partial autocorrelation functions for the residuals from this regression. Interpret them.

````{r}

acf(resid(mod2))
pacf(resid(mod2))

````

5) Fit an autoregressive models to the residuals. Compare the results with your interpretation of the PACF plot.

````{r}
ar(x = resid(mod2))

````


6) Fit a gls model using a corAR1 correlation structure. Test if the correlation structure is necessary. Is the trend significant? 

````{r}
year <- time(LakeHuron)
fit3.gls <- gls(LakeHuron~year)
fit4.gls <- gls(LakeHuron~year, corr = corAR1())

anova(fit3.gls, fit4.gls)

````

7) Fit a gls model using a corARMA correlation structure with two AR terms. Is this model an improvement?

````{r}
fit5.gls <- gls(LakeHuron~year, corr = corARMA(p=1, q=0)) #p is the number of AR terms, q is the number of MA terms. So to fit an AR2 model you need p = 2, q = 0

anova(fit3.gls, fit4.gls, fit5.gls)

````
